Persistence in functional topology: local-connectivity, q-tameness and a correction to a theorem of Morse

### 2021-05-14 surfaces.tex
Updates from Overleaf
Modify Section: surfaces, bring Morse discussion there

Throughout his work on functional topology, to obtain q-tameness Morse assumed slightly varying forms of local-connectivity on the resulting sublevel set filtrations.
In particular, Morse and Tompkins use in their applications to minimal surface theory the following version:
\begin{displaycquote}[p.431]{Morse.1940}
	The space $M$ is said to be \textit{locally $F$-connected} of order $r$ at $p$ if corresponding to each positive constant $e$ there exists a positive constant $\delta$ such that each singular $r$-sphere on the $\delta$-neighborhood of $p$ on $F_{c+\delta}$ bounds an $(r+1)$-cell of norm $e$ on $F_{c+e}$.
\end{displaycquote}
Here, $M$ is a metric space and $F$ a real-valued function on $M$ whose sublevel sets $F_{t}$ are all compact.
Morse then states that as consequence of these assumptions the persistent \v{C}ech homology of this sublevel set filtration is q-tame.
In the original it reads:
\begin{displaycquote}[Theorem 6.3, p.432]{Morse.1940}
	Let $a$ and $c$ be positive constants such that $a < c < 1$.
	The $k^{\mathrm{th}}$ connectivity $R^k(a,c)$ of $F_a$ on $F_c$ is finite.
\end{displaycquote}
Morse does not prove this statement in the given reference, but rather refers to \cite[Theorem 6.1]{Morse.1938}.
Unfortunately, the claim as stated is not correct and in this work we provide a counterexample as a consequence of \cref{thm:counterexample}.

@@

Morse and Tompkins prove that each homotopically critical point of the functional $A_g$ corresponds to a critical point of the area functional, i.e., a minimal surface.
They then use a form of Morse inequalities for homotopically critical points
% , or, more generally, critical sets
to infer their Unstable Minimal Surface Theorem:
\begin{itemize}
    \item[($\ast$)] If $\Omega_g$ contains two distinct solutions to Plateau's Problem then it also contains an unstable minimal surface, more precisely a critical point of index 1.
\end{itemize}

In \cref{s:inequalities} we review Morse's topological inequalities using the language of barcodes.
In order to apply these inequalities and deduce their theorem, Morse and Tompkins need to prove the \mbox{q-tameness} of $(\Omega_g, A_g)$. 
For this they want to use that $(\Omega_g, A_g)$ is weakly $\piLC$ \cite[p.431]{Morse.1940}. This property is not sufficient to ensure q-tameness as \cref{thm:counterexample} demonstrates.
However, the proof given in \cite[p.464]{Morse.1939} establishes that $(\Omega_g, A_g)$ is $\LC$, implying that it is also $\HLC$ for any homology theory.
Therefore, from \cref{t:strong local connectedness implies q-tameness} we can conclude that $(\Omega_g, A_g)$ is q-tame as claimed.
This implies that Morse inequalities hold as does the unstable minimal surfaces theorem of Morse.

% We also mention that in \cite[p.421]{Morse.1937} appearing three years before the work of Morse-Tompkins on minimal surfaces, Morse considered the notion of $\piLC$ and claiming it to be sufficient to ensure q-tameness of compact sublevel set filtrations.

% We anticipate that the stronger version of local $F$-connectivity introduced by Morse, and inspiring our $\HLC$ notion, is also sufficient to obtain q-tameness, but we do not provide a proof.

% A stronger notion of local connectivity for the pair $(M, F)$ appeared in print, introduced by Morse himself three years earlier, and claimed sufficient for q-tameness although no proof was given.
% \begin{displaycquote}[p.421]{Morse.1937}
% 	The space $M$ will be said to be locally $F$-connected for the order $n$ if corresponding to $n$, an arbitrary point $p$ on $M$, and an arbitrary positive constant $e$, there exists a positive constant $\delta$ with the following property. For $c \geq F(p)$ any singular $n$-sphere on $F \leq c$ (the continuous image on $F \leq c$ of an ordinary $n$-sphere) on the $\delta$-neighborhood $p_{\delta}$ of $p$ is the boundary of a singular $(n + 1)$-cell on $F \leq c + e$ and on $p_e$.
% \end{displaycquote}
% This is precisely the notions of $\piLC$ \cref{defi:local_connectedness_filtrations}.

% As is, the argument given by Morse and Tompkins to establish $(\ast)$ is not strictly valid, however, the proof of the local $A_g$-connectivity of $\Omega_g$ given in \cite[p.?]{Morse.1939} establishes th stronger statement implying, in our terminology, that $(\Omega_g, A_g)$ is $\HLC$ for any homology theory, and \cref{t:strong local connectedness implies q-tameness} fixes this gap.
% We anticipate that the stronger version of local $F$-connectivity introduced by Morse, and inspiring our $\HLC$ notion, is also sufficient to obtain q-tameness, but we do not provide a proof.

>>>>

@@

Morse and Tompkins prove that each homotopically critical point of the functional $A_g$ corresponds to a critical point of the area functional, i.e., a minimal surface.
They then use a form of Morse inequalities for homotopically critical points to infer their Unstable Minimal Surface Theorem:
\begin{itemize}
    \item[($\ast$)] If $\Omega_g$ contains two distinct solutions to Plateau's Problem then it also contains an unstable minimal surface, more precisely a critical point of index 1.
\end{itemize}

In order to apply these inequalities and deduce their theorem, Morse and Tompkins need to prove the \mbox{q-tameness} of $(\Omega_g, A_g)$.
Throughout his work on functional topology, to obtain q-tameness Morse assumed slightly varying forms of local-connectivity on the resulting sublevel set filtrations.
In particular, Morse and Tompkins use in their applications to minimal surface theory the following version equivalent to $\piLC$:
\begin{displaycquote}[p.431]{Morse.1940}
	The space $M$ is said to be \textit{locally $F$-connected} of order $r$ at $p$ if corresponding to each positive constant $e$ there exists a positive constant $\delta$ such that each singular $r$-sphere on the $\delta$-neighborhood of $p$ on $F_{c+\delta}$ bounds an $(r+1)$-cell of norm $e$ on $F_{c+e}$.
\end{displaycquote}
They then state that as a consequence of these assumptions the persistent \v{C}ech homology of this sublevel set filtration is q-tame.
In the original it reads:
\begin{displaycquote}[Theorem 6.3, p.432]{Morse.1940}
	Let $a$ and $c$ be positive constants such that $a < c < 1$.
	The $k^{\mathrm{th}}$ connectivity $R^k(a,c)$ of $F_a$ on $F_c$ is finite.
\end{displaycquote}
Morse does not prove this statement in the given reference, but rather refers to \cite[Theorem~6.1]{Morse.1938}.
Unfortunately, this argument is not in general sound as \cref{thm:counterexample} shows.
However, the proof given in \cite[p.464]{Morse.1939} establishes that $(\Omega_g, A_g)$ is $\LC$, implying that it is also $\HLC$ for any homology theory.
Therefore, from \cref{t:strong local connectedness implies q-tameness} we can conclude that $(\Omega_g, A_g)$ is q-tame as claimed.
This implies that Morse inequalities hold, and hence so does the unstable minimal surfaces theorem of Morse.
